\section{Implementation}
\label{sec:impl}

\subsection{Class Declaration Processing}

When \texttt{voo::class} is evaluated, VOO creates a namespace, parses field and method
declarations, assigns sequential field indices, and generates all accessors and
constructors. For inherited classes, parent field indices and defaults are copied first,
and new fields continue sequentially. According to experiments, declaring 1,000 small classes (5 fields, single method) using VOO costs approximately 300~\textmu s
versus 50~\textmu s for TclOO---an acceptable one-time cost during application
initialization, given the substantial runtime benefits.

\subsection{Copy-on-Write Optimization}

VOO's updater pattern avoids COW propagation during nested modifications by temporarily
extracting the field value into a local variable, setting the field slot to empty
(detaching it), executing the user's modification script, and reattaching the value in
a \texttt{finally} block. When the local variable holds the sole reference, modifications
proceed without copying the parent object.

\subsection{Migration Path to C++}

VOO's architecture enables seamless migration to compiled C++ via a companion
header-only template framework organized around three responsibilities.
\texttt{tcl::Type<T>} registers a custom \texttt{Tcl\_ObjType} for any C++ class,
providing COW semantics and reference counting through standard callbacks; only
\texttt{TypeName()} must be specialized. \texttt{tcl::obj\_cast} provides bidirectional
type casting between \texttt{Tcl\_Obj*} and C++ types with built-in specializations for
fundamental types. \texttt{TCLCPPG\_*} macros generate \texttt{Tcl\_CreateObjCommand}
registrations with lambda wrappers, argument validation, automatic type conversion, and
error handling.

The migration follows four steps: define a normal C++ class, specialize
\texttt{tcl::Type<T>}, register commands using macros, and use from Tcl---with the call
site identical to the VOO Tcl API\@. Migration requires no changes to caller code; only
the class declaration is replaced and commands re-registered by the C++ package
initializer.
